---
title: "The Binomial Distribution"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Binomial Distribution}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The binomial distribution is the key to exact inference about a proportion. 
The purpose of this vignette is to explain the binomial distribution and then 
demonstrate how it can be used for exact inference, including both hypothesis 
testing and confidence interval construction.

### Required packages

The nplearn package is required for this vignette. All other functions used 
are in the base package.

```{r setup, echo = TRUE, message = FALSE}
library(nplearn)
```

### Joint Events

Multiple independent events that occur are referred to as *joint events*. 
The total number of possible joint outcomes of a joint event is the product 
of the number of potential outcomes of each separate event. For example, on 
a five-item multiple choice quiz, where each question has four options, the 
number of possible patterns of responses is the product of five fours.

```{r}
4*4*4*4*4
```

We can write this as an exponent.

```{r}
4^5
```

We know from the laws of probability that if each event is independent, the probability of a specific set of joint events is equal to the product of the 
probability of the individual events. In our multiple-choice quiz example, 
here is the probability of a test taker getting a perfect score just by 
guessing.

```{r}
(1/4)*(1/4)*(1/4)*(1/4)*(1/4)
```

Here it is written another way.

```{r}
(1/4)^5
```

Here it is written yet another way.

```{r}
0.25^5
```

Suppose we have a series of Bernoulli events. For example, suppose that we 
again administer a five-question quiz, but this time the items are true-false. 
This is a series of Bernoulli events because there are only two possible 
outcomes for each event. Here is the probability of getting a perfect score 
just by guessing.

```{r}
0.5^5
```

A more sophisticated test taker will actually have studied for the test, so 
this test taker will not just be guessing. We could think of the probability 
of success on an item as a function of this person's knowledge after learning 
the content. Here is the probability of a perfect score if the probability of 
success on each item is 0.7.

```{r}
0.7^5
```

Let's see what the chance of a perfect score is for someone who is very well 
prepared so that the success rate for each item is 0.95.

```{r}
0.95^5
```

Of course all of this is making the unrealistic assumption that the rate of 
success is the same for every item. In fact, so items are more difficult than 
others, so the actual probability of a perfect score might be something more 
like this.

```{r}
0.8*0.95*0.75*0.95*0.70
```

This kind of test-taking discussion with different item difficulties is best 
left for a measurement course, such as item response theory. We will step 
out of the test-taking domain in a bit, but for now let's proceed with this 
example and the (unrealistic) assumption that each item has the same 
difficulty level.

### Permutations

Suppose I'm not just interested in a perfect score, but I'm also interested 
in questions like, what is the chance of obtaining four out of five correct 
on this true-false quiz if the probability of success for each item is 0.7? 
This is a bit tricky, because we don't know which item the person will get 
wrong. Here is the probability of getting just the first item wrong.

```{r}
0.3*0.7*0.7*0.7*0.7
```

Here it is written another way.

```{r}
0.7^4 * 0.3^1
```

What about the probability of getting just the second item wrong?

```{r}
0.7*0.3*0.7*0.7*0.7
```

The probability is the same! In fact, we could still write it this way.

```{r}
0.7^4 * 0.3^1
```

It is pretty easy to figure out that there are five possible ways for the 
test taker to get just one item wrong on a five-item test: The first one could 
be wrong, the second item could be wrong, and so on. Thus, the total 
probability of getting any one item wrong is calculated this way.

```{r}
5 * 0.7^4 * 0.3^1
```

That was straightforward, but it becomes more difficult with different 
numbers of items correct and different numbers of items. For example, how many 
different patterns will there be for three items correct and two items 
incorrect? What about if there are 10 items instead of five? How many patterns 
will there be for four items correct and six incorrect? Clearly these 
questions are not as easy to answer. What we need are some counting rules. 
Many nonparametric statistics are based on working with discrete values, 
so we often need to count. Methods of counting for various situations are 
an important component of nonparametric analysis.

The first counting rule is about permutations. A permutation of objects is a 
distinct ordering of objects. We can calculate the number of permutations 
of N objects with a product known as a *factorial*. For example, for five 
objects we use 5! (read as, five factorial).

```{r}
5 * 4 * 3 * 2 * 1
```

R includes a factorial function.

```{r}
factorial(5)
```

A side note: What do we do with 0!? By definition, 0! = 1.

```{r}
factorial(0)
```

### The Binomial Coefficient

In our true-false quiz example we had five questions, so the number of possible 
permutations of the answers is 5! = 120. Yet if we want to know how many ways 
there are to get two incorrect answers out of the five questions, we really 
only care about two incorrect and three correct answers, regardless of where 
they occur. That is, we consider all incorrect answers as indisitinguishable 
from one another. There are only two distinct type of answers here: correct 
and incorrect. The permutation of wrong, wrong, right, right, right does not 
need to be considered different than the permutation of wrong, right, right, 
right, wrong.

To calculate how many permutations there are when we include identical 
"objects" (we consider every "right" and every "wrong" as identical) we need 
to divide the total number of permutations by the number of permutations of 
each identical object. Here's the calculation for five true-false items when 
three are correct and two are incorrect.

```{r}
factorial(5)/(factorial(3)*factorial(2))
```

This can be extended to any number of identical objects. The case where we 
have only two identical objects is all we will need when looking at a series 
of Bernoulli events, because by definition a Bernoulli event only has one of 
two outcomes. Given this case is so special, there is a special expression for 
it, as well as a special function for it. When we want to know how many ways 
we can get three items correct out of five Bernoulli trials, we say 
*five choose three*. Here is how we represent it.

$\binom{5}{3}$

Here's the R function.

```{r}
choose(5, 3)
```

This function is referred to as the *binomial coefficient*. In general, we 
represent the number of ways that we can have $x$ successes in $n$ trials 
like this, which we read as *n choose x*.

$\binom{n}{x}$

It would seem quite difficult to figure how many different ways there are 
to obtain 13 correct answers on a 20-item true-false quiz, but the 
binomial coefficient function makes it easy.

```{r}
choose(20,13)
```

### The Binomial Probability Function

If we combine the concepts of joint events with the binomial coefficient, 
we can answer an interesting question. When we observe $n$ independent Bernoulli 
events, each with the same probability of success, $pi$, what is the 
probability that $x$ of these events will be a success? If that doesn't really 
sound very interesting to you, consider a couple of examples.

If the overall pass rate on an entrance exam is 70%, what is the probability 
that we will be able to fill eight slots with successful examinees if there 
are 11 applicants for the position?

If the chance of contracting an illness is 10% when coming into contact with 
another person with the illness, what is the probability that all six people 
who came into contact with an ill person will avoid aquiring the illness 
themself?

As we will see shortly, knowing the probability of $x$ successes in $n$ trials 
can also be used to conduct hypotheses tests and construct confidence intervals 
for a proportion, so the value of being able to calculate probabilities based 
on $n$ Bernoulli events goes beyond answering questions like those above.

Recall that we calculate the probability of independent joint events by 
taking the product of the probability of each event. When there is a 70% chance 
of passing an item on a five-item quiz, we calculate the probability of 
getting the first three items correct but missing the two remaining items like 
this.

```{r}
0.7^3 * 0.3^2
```

In more general terms, the chance of getting the first $x$ items correct but 
missing the remaining items on an $n$-item quiz when the probability of 
success is $pi$ is calculated like this.

$\pi^x (1-\pi)^{(n-x)}$

That is, we multiply the $x$ successes by the $n-x$ failures.

I can't think of a reason why I would want to know the chance of getting the 
first three items correct and missing the last two, but I can think of why 
I might want to know the probability of missing *any* two, not just the last 
two. That is where the binomial coefficient comes in handy. The binomial 
coefficient tells us how many ways we can arrange $x$ successes in $n$ trials. 
Here is how many ways we can have exactly three items correct on a five-item 
quiz.

```{r}
choose(5, 3)
```

If I multiply the probability of one specific sequence of three correct and 
two wrong with the number of ways I can have three correct and two wrong, that 
will give me the probability of obtaining any three correct on a five-item 
quiz.

```{r}
choose(5, 3) * 0.7^3 * 0.3^2
```

We can right this in general terms like this.

$P(X = x) = \binom{n}{x}\pi^x (1-\pi)^{(n-x)}$

This is the binomial probability function. It tells us the probability of 
$x$ successes in $n$ Bernoulli events when the probability of success is $\pi$ 
for each Bernoulli event. Note that whereas the values of $x$ are either 0 or 
1 for a Bernoulli event, the values of $x$ for the binomial function will 
range from 0 to $n$. That is, you can have 0 success, 1 success, 2 successes, 
and so on, all the way up to $n$ successes.

Let's now answer one of our two questions. If the chance of contracting an 
illness is 10% when coming into contact with another person with the illness, 
what is the probability that all six people who came into contact with an ill 
person will avoid aquiring the illness themself? This is 
