---
title: "Workflow Sample"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Workflow Sample}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The purpose of this vignette is to illustrate the typical steps and assumptions 
used by researchers when analyzing data. The illustration uses the data from a 
study of twin intelligence. In 1966, Cyril Burt published data obtained from 27 
pairs of monozygotic twins. These 27 pairs were special cases in that one twin 
was raised in a foster home, while the other was raised by the birth parents. 
Burt measured the intelligence of the adult twins in order to determine whether environmental factors influence intelligence, or whether it is primarily a 
function of genetics. He also classified each twin pair as having come from a 
“low,” “medium,” or “high” social class.

### Reference

Burt, C. (1966). The genetic determination of differences in intelligence: A 
study of monozygotic twins reared together and apart. *British Journal of 
Psychology, 57,* 137 –153.

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

data_file <- system.file("extdata", "IQ_Study_Data.csv", package = "nplearn")
```

### Required packages

This example uses packages that are part of the tidyverse. Installation of 
the complete tidyverse package will facilitate running this script.

```{r setup, echo = TRUE, message = FALSE}
library(nplearn)
library(tidyverse)
```

### Data Import

First, data are imported into the statistical software.

```{r message = FALSE}
iq_data <- read_csv(data_file)
```

### Data Cleaning and Preparation

Next, we clean data and prepare it for analysis. In this case, because we 
have repeated measurements (due to correlated twin scores) in the original 
data set, we will need both long and wide formats. The data are currently in 
wide format only. We also should obtain a variable for the difference in 
IQ scores for each twin pair.

```{r}
iq_data_long <- gather(iq_data,
                       'Foster',
                       'Biological',
                       key = "Upbringing",
                       value = "IQ")

iq_data$IQ_Diff <- iq_data$Foster - iq_data$Biological

iq_data$Social <- as.factor(iq_data$Social)
iq_data_long$Social <- as.factor(iq_data_long$Social)

iq_data$Social <- fct_relevel(iq_data$Social,
                              c("low", "middle", "high"))

iq_data_long$Social <- fct_relevel(iq_data_long$Social,
                                   c("low", "middle", "high"))
```

### Graphical Display

Next, we construct graphical displays to illustrate the major features of the 
data. First, side-by-side boxplots to compare the foster twins IQ to the 
biological twins IQ.

```{r fig1, fig.height = 5, fig.width = 7, fig.align = "center"}
ggplot(iq_data_long, aes(x = Upbringing, y = IQ)) +
  geom_boxplot() +
  labs(x = "Type of Upbringing",
       y = "Intelligence Quotient",
       title = "Comparison of IQ Scores for Twins with Different Upbringing")
```

Here are boxplots to compare IQ score differences for each twin pair across 
the three SES categories.

```{r fig2, fig.height = 5, fig.width = 7, fig.align = "center"}
ggplot(iq_data, aes(x = Social, y = IQ_Diff)) +
  geom_boxplot() +
  labs(x = "Socioeconomic Status",
       y = "IQ Difference Scores",
       title = "Comparison of Twin IQ Differences for SES Categories")
```

Here is a scatterplot illustrating the relationship of IQ scores for foster 
home and biological twins. A regression line is included, as well as 
differing colors and shapes for the social categories. Social category does 
not appear to make much difference in the relationship, so a single regression 
line is sufficient.

```{r fig3, fig.height = 5, fig.width = 7, fig.align = "center"}
ggplot(iq_data, aes(x = Biological, y = Foster)) +
  geom_point(aes(color = Social, shape = Social)) +
  geom_smooth(method = lm, se = FALSE) +
  labs(x = "Biological Twin IQ Score",
       y = "Foster Home Twin IQ Score",
       title = "Relationship of Twin IQ Scores")
```

### Descriptive Statistics

Descriptive statistics are indices of characteristics of a set of scores. We 
can calculate these for the entire set of scores, or for subsets. These should 
align with our graphs and provide more details.

Here are statistics comparing IQ scores for different types of upbringing.

```{r}
iq_data_long %>%
  group_by(Upbringing) %>%
  summarize(mean(IQ),
            sd(IQ),
            min(IQ),
            quantile(IQ, 0.25),
            median(IQ),
            quantile(IQ, 0.75),
            max(IQ))
```

Here are statistics comparing IQ difference scores for different social 
categories.

```{r}
iq_data %>%
  group_by(Social) %>%
  summarize(mean(IQ_Diff),
            sd(IQ_Diff),
            min(IQ_Diff),
            quantile(IQ_Diff, 0.25),
            median(IQ_Diff),
            quantile(IQ_Diff, 0.75),
            max(IQ_Diff))
```

Here are the coefficients for the regression line in the model relating 
the foster home twin IQ score to the biological twin IQ score, as well as the 
correlation among these scores.

```{r}
lm(iq_data$Foster ~ iq_data$Biological)
cor(iq_data$Foster, iq_data$Biological)
```

### Inferential Statistics

Inferential statistics attach a degree of probability to point estimates that 
are compared to hypothesized values or to interval estimates. Such statistics 
are a function of effect size, variation, and form of the population 
distribution. Inferences should not be made in isolation, but only to 
supplement graphical and descriptive analysis.

This generates a t-test confidence interval for the comparison of foster home 
twin IQ score to biological twin IQ score.

```{r}
t.test(iq_data$Foster, iq_data$Biological)
```

This provides a *p* value for the test that all social categories have the 
same mean IQ score. It is followed by Tukey confidence intervals, to control 
the overall Type I error rate.

```{r}
anova(lm(iq_data$IQ_Diff ~ iq_data$Social))

TukeyHSD(aov(iq_data$IQ_Diff ~ iq_data$Social))
```

This provides a *p* value for the test that there is no relationshiop of the 
foster twin IQ score to the biological twin IQ score. It is followed by 
the construction of a confidence interval for the correlation among these 
variables.

```{r}
anova(lm(iq_data$Foster ~ iq_data$Biological))

cor.test(iq_data$Foster, iq_data$Biological)
```
