---
title: "Bernoulli Events"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bernoulli-Events}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Bernoulli events have just two potential outcomes, often referred to as 
"success" or "failure". (You need to consider these terms loosely. For example, 
if working with polling data, one candidate would consider voting a certain way 
as a success whereas the other candidate would certainly consider it a 
failure.) 

### Required packages

This vignette uses the nplearn and Rlab packages.

```{r setup, echo = TRUE, message = FALSE}
library(nplearn)
library(Rlab)
```

### Probability distributions

A Bernoulli probability distribution is the simplest type of probability distribution.

Here is a Bernoulli distribution for $\pi$ = 0.7.

```{r}
bern_plot(0.7)
```

A probability distribution based on a discrete event is the same as a density 
distribution. That is, the density and the probability are identical. Here is 
the density for x = 0 (a failure) when $\pi$ = 0.7. Note that it is identical 
to the probability of a failure.

```{r}
dbern(0, 0.7)
```

This is different than the probability and density with continous 
distributions. The probability of any single value of a continuous distribution 
is 0. When we look at a plot for a continuous distribution, we are looking at 
a density plot, with potential values of the variable as the horizontal 
dimension and density as the vertical dimension. Here's the density plot for 
the standard normal distribution. Note the function "dnorm" in the code. This 
is short for "density for the normal distribution."

```{r}
s <- seq(-4,4,.01)

plot(s, dnorm(s), type="l", lwd = 3, ylim = c(0,0.4),
     xlab = "Z Scores", ylab = "Density", main = "Standard Normal Distribution")

```

### Cumulative probability

In order to find probability for values on a continuous distribution, we must 
take a range of values in a continuous distribution. For example, here is the  probability of obtaining a value between 0.2 and 0.3 on the standard normal  distribution.

```{r}
pnorm(0.3) - pnorm(0.2)
```

The pnorm function provides us the "cumulative probability" up to the 
specified value. Thus, the "pnorm(0.3)" gives us the area on the density curve 
that is below 0.3.

```{r}
pnorm(0.3)

s <- seq(-4,4,.01)

plot(s, dnorm(s), type="l", lwd = 3, ylim = c(0,0.4),
     xlab = "Z Scores", ylab = "Density", main = "Standard Normal Distribution")

segments(x0 = 0.3, y0 = 0, x1 = 0.3, y1 = dnorm(0.3))
```

By subtracting the cumulative probability of one score from a second score, 
we can obtain the probability of obtaining a score between these two scores.

```{r}
pnorm(0.3)
pnorm(0.2)
pnorm(0.3) - pnorm(0.2)

s <- seq(-4,4,.01)

plot(s, dnorm(s), type="l", lwd = 3, ylim = c(0,0.4),
     xlab = "Z Scores", ylab = "Density", main = "Standard Normal Distribution")

segments(x0 = 0.2, y0 = 0, x1 = 0.2, y1 = dnorm(0.2))
segments(x0 = 0.3, y0 = 0, x1 = 0.3, y1 = dnorm(0.3))
```

When we run a cumulative probability function for a continuous distribution,
behind the scenes the function is using numerical integration to calculate 
the area under the curve. This is something you learn about in a calculus 
class. One reason that nonparametric statistical analysis can be considerably 
easier than traditional parametric analysis is that we work with discrete 
values, so we can do probability calculations with simple arithmetic since 
the probability is the same as the density.

Here's the cumulative probability for x = 1 on the Bernoulli distribution.

```{r}
pbern(1, 0.7)
```

The cumulative probability for x = 1 is found by summing the individual 
probabilities for x = 0 and x = 1. In short, this tells us that when the only 
outcomes possible are a success and a failure, the chance of obtaining one 
of these in a Bernoulli event is 100%.

```{r}
dbern(0, 0.7) + dbern(1, 0.7)
```

### Quantiles

The general term for a score in a distribution (pictured on the horizontal 
axis for a graph of a density distribution) is "quantile". Quantile functions 
are the opposite of cumulative probability functions. We submit the 
cumulative probability and the function will tell us what quantile has that 
cumulative probability. Recall that the cumulative probability for 0.3 on the 
standard normal distribution is 0.6179114.

```{r}
qnorm(0.6179114)
```

Here is the quantile for the cumulative probability of 0.3 on a Bernoulli 
distribution with a probility of success of 0.7.

```{r}
qbern(0.3, 0.7)
```

Here is the quantile for the cumulative probability of 0.7 on a Bernoulli 
distribution with a probility of success of 0.7.

```{r}
qbern(0.7, 0.7)
```

What is the quantile for the cumulative probability of 0.5 on a Bernoulli 
distribution with a probility of success of 0.7. This can be confusing 
because there are only two quantiles on the Bernoulli distribution. One 
characteristic of discrete distributions is that not every cumulative 
probability relates in a one-to-one fashion with a quantile, like it does on 
a continuous distribution. Quantiles are often defined as the smallest value of 
x on the distribution such that the cumulative probability to x is at least 
the specified probability. For example, if we want the quantile corresponding 
to a cumulative probability of 0.5 on the Bernoulli distribution, we need to 
find the smallest quantile that has a cumulative probability of at least 0.5.

```{r}
qbern(0.5, 0.7)
```

In R there are functions for density (d), cumulative probability (p), 
and quantiles (q) for many distributions. We have looked at dnorm, pnorm, 
qnorm, dbern, pbern, and qbern. By default, the p and q functions are based on  accumulating from the lower tail, but we can switch that to the upper tail by  setting lower.tail = FALSE.

```{r}
1 - pnorm(0.3)
pnorm(0.3, lower.tail = FALSE)

1 - pbern(0, 0.7)
pbern(0, 0.7, lower.tail = FALSE)
```

### Random variates

R also has (r) functions for many distributions (e.g., rnorm and 
rbern). These provide us a specified number of randomly seleted values from 
the corresponding distribution. These are known as "random variates". This will 
be very handy when we study concepts such as Monte Carlo methods.

```{r}
rnorm(10)
rbern(10, 0.7)
```

