---
title: "Workflow Sample"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Workflow Sample}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The purpose of this vignette is to illustrate the typical steps and assumptions 
used by researchers when analyzing data. The illustration uses the data from a 
study of twin intelligence. In 1966, Cyril Burt published data obtained from 27 
pairs of monozygotic twins. These 27 pairs were special cases in that one twin 
was raised in a foster home, while the other was raised by the birth parents. 
Burt measured the intelligence of the adult twins in order to determine whether environmental factors influence intelligence, or whether it is primarily a 
function of genetics. He also classified each twin pair as having come from a 
“low,” “medium,” or “high” social class.

### Reference

Burt, C. (1966). The genetic determination of differences in intelligence: A 
study of monozygotic twins reared together and apart. *British Journal of 
Psychology, 57,* 137 –153.

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

data_file <- system.file("extdata", "IQ_Study_Data.csv", package = "nplearn")
```

### Required packages

This example uses packages that are part of the tidyverse. Installation of 
the complete tidyverse package will facilitate running this script.

```{r setup, echo = FALSE}
library(nplearn)
library(tidyverse)
```

### Data Import

First, data are imported into the statistical software.

```{r}
iq_data <- read_csv(data_file)
```

### Data Cleaning and Preparation

Next, we clean data and prepare it for analysis. In this case, because we 
have repeated measurements (due to correlated twin scores) in the original 
data set, we will need both long and wide formats. The data are currently in 
wide format only. We also should obtain a variable for the difference in 
IQ scores for each twin pair.

```{r}
iq_data_long <- gather(iq_data,
                       'Foster',
                       'Biological',
                       key = "Upbringing",
                       value = "IQ")

iq_data$IQ_Diff <- iq_data$Foster - iq_data$Biological

iq_data$Social <- as.factor(iq_data$Social)
iq_data_long$Social <- as.factor(iq_data_long$Social)

iq_data$Social <- fct_relevel(iq_data$Social,
                              c("low", "middle", "high"))

iq_data_long$Social <- fct_relevel(iq_data_long$Social,
                                   c("low", "middle", "high"))
```

### Graphical Display

Next, we construct graphical displays to illustrate the major features of the 
data.

```{r}
ggplot(iq_data_long, aes(x = Upbringing, y = IQ)) +
  geom_boxplot()

ggplot(iq_data, aes(x = Social, y = IQ_Diff)) +
  geom_boxplot()

ggplot(iq_data, aes(x = Biological, y = Foster)) +
  geom_point(aes(color = Social, shape = Social)) +
  geom_smooth(method = lm, se = FALSE)
```

### Descriptive Statistics

Descriptive statistics are indices of characteristics of a set of scores. We 
can calculate these for the entire set of scores, or for subsets. These should 
align with our graphs and provide more details.

```{r}
iq_data_long %>%
  group_by(Upbringing) %>%
  summarize(mean(IQ),
            sd(IQ),
            min(IQ),
            quantile(IQ, 0.25),
            median(IQ),
            quantile(IQ, 0.75),
            max(IQ))

iq_data %>%
  group_by(Social) %>%
  summarize(mean(IQ_Diff),
            sd(IQ_Diff),
            min(IQ_Diff),
            quantile(IQ_Diff, 0.25),
            median(IQ_Diff),
            quantile(IQ_Diff, 0.75),
            max(IQ_Diff))

lm(iq_data$Foster ~ iq_data$Biological)
cor(iq_data$Foster, iq_data$Biological)
```

### Inferential Statistics

Inferential statistics attach a degree of probability to point estimates that 
are compared to hypothesized values or to interval estimates. Such statistics 
are a function of effect size, variation, and form of the population 
distribution. Inferences should not be made in isolation, but only to 
supplement graphical and descriptive analysis.

```{r}
t.test(iq_data$Foster, iq_data$Biological)

anova(lm(iq_data$IQ_Diff ~ iq_data$Social))

TukeyHSD(aov(iq_data$IQ_Diff ~ iq_data$Social))

anova(lm(iq_data$Foster ~ iq_data$Biological))

cor.test(iq_data$Foster, iq_data$Biological)
```

